Some forms of uncertainty may suppress the evolution of social learning

### 2022-04-19 ms.tex
finished dynamics and evolution subsec first draft

\newcommand{\pisub}[1]{\pi_{\mathrm{#1}}}
\newcommand{\pilow}{\pisub{low}}

@@

In each time step, $N$ agents---autonomous problem solvers---select 
which behavior to perform based on their running tally of mean payoffs for 
each available behavior. %information (Figure~\ref{fig:behaviorSelection}). 
Agents track the mean payoff of each behavior $b$, denoted $\bar\pi_b$, and a \emph{count} of how many times it has performed each behavior, denoted $c_b$. 
Agents' beliefs about mean payoffs are updated from $\bar\pi_b$ to $\bar\pi_b'$ using exponential weighted
averaging, $\bar\pi_b' = \bar\pi_b + \frac{\mathrm{Bandit_b(0, 1)} - \bar\pi_b}{c_b'}$, where
$\mathrm{Bandit}_b(0, 1)$ is 0 or 1 depending on the result of the bandit draw for behavior $b$. 
$\bar\pi_b$ is initialized to 0 for all $b$ at model initialization for
all agents, and after each generation for individual learner agents. Social
learning agents' $\bar\pi_b$ is initialized as that of its teacher selected
through performance-biased oblique learning.

@@

Following Luce's choice axiom \cite{Luce1959}, an agent using asocial information selects a behavior at random, with the probability of selecting any particular behavior weighted by the softmax function applied to that behavior's observed mean payoff relative to all 
mean payoffs: 
$\Pr(\text{choose behavior }b) = 
  \frac{\exp(\bar\pi_b / \tau) }{ 
\sum_{b=1}^B \exp(\bar\pi_b / \tau)}$.
The softmax method is a biologically plausible model~\cite{Schulz2019} that enables agents to explore alternative behaviors sometimes and exploit the best observed behavior other times.  The parameter $\tau$ specifies
how frequently alternative behaviors are explored---higher $\tau$ means more
frequent exploration. We set $\tau = 0.1$ for all computational analyses 
presented in the main paper. We (WILL HAVE) performed sensitivity analyses 
and found a weak dependence on $\tau$ that does not affect our main conclusions.

@@

In our model uncertainty is a tunable consequence of four environmental features:
%that structure the behavior of agents.  
% Many factors contribute to the 
% uncertainty involved in both social and asocial learning. 
%We model uncertainty in four distinct ways: 
\begin{enumerate}
    \item We vary the frequency at which the optimal behavior changes, which we call \emph{environmental variability}, $u$. At the start of each generation, with probability $u$, a new behavior is assigned a payoff of $\pisub{high}$ while all other behaviors are assigned a payoff of $\pisub{low}$. Otherwise, the same behavior remains optimal across generations. 
    \item We vary the latent expected payoffs yielded by the bandits. For simplicity,
      we assume that in any given environmental state, there is one optimal behavior that
      yields an expected payoff of $\pisub{high}$, while all other behaviors yield a 
      payoff of $\pisub{low} < \pisub{high}$. 
      The difference between these quantities is the \emph{payoff ambiguity}, 
    \begin{equation}
      A_\pi = 1 - (\pisub{high} - \pisub{low}),
      \label{eq:payoffAmbiguity}
    \end{equation}
    which is maximized when the expected payoffs are close to equal and minimized
    when they differ greatly. For simplicity we will set $\pisub{high} = 0.9$
    for all analyses and vary $\pisub{low}$ in order to vary $A_\pi$.
    \item We vary the total \emph{number of available behaviors}, $B$, which is a
      source of uncertainty since agents are less likely to know which behavior yields high payoffs.  %risk, as it becomes increasingly costly to experiment with alternatives. \cm{is this risk if it reduces variance of outcomes? it sounds like it's just bad for there to be more 0 payoff options}
    \item Finally, we vary number of behavioral events per generation, $M$. This can be viewed as the \emph{effective lifespan} of an agent.  Decreasing this lifespan effectively increases the importance of each event for acquiring payoffs. When lifespans are short ($M$ is small), agents experience greater uncertainty about which behavior is optimal given the fewer learning opportunities.
\end{enumerate}

@@

@@

@@

@@

  \item 
    However, more possible behaviors $B$ ($B=2,4,10$ for the first, second, and
    third column respectively in Figure~\ref{fig:results}) can lead
    social learning to be selected for even for large values of $u$, possibly
    due to the low risk and potentially high reward of learning an accurate
    tally of mean payoffs from the previous generation. 
    Since there are many behaviors and
    few observations, softmax weighting will not significantly 
    favor a sub-optimal behavior. However, if one does learn an accurate tally
    then one does have an increased chance of performing the best one.
    \begin{itemize}
      \item 
        \mt{This makes sense, but the effect is too strong. Why should there be any
        social learning when u = 1.0?  I think something is slightly wrong with the
        model---I'm working on checking what that might be}
      \item
        \mt{One thing to check is the fraction doing the optimal behavior when
        $u=1$. If many start out doing optimal, then something may be wrong.}
    \end{itemize}

>>>>

\newcommand{\pisub}[1]{\pi_{\mathrm{#1}}}
\newcommand{\pilow}{\pisub{low}}
\newcommand{\pihigh}{\pisub{high}}

@@

In each time step, $N$ agents---autonomous problem solvers---select 
which behavior to perform based on their running tally of mean payoffs for 
each available behavior. %information (Figure~\ref{fig:behaviorSelection}). 
Agent $i$ tracks the mean payoff of each behavior $b$, denoted $\bar\pi_{ib}$, and
a \emph{count} of how many times it has performed each behavior, denoted $c_{ib}$.
$\bar\pi_{ib}$
is initialized to 0 for all $b$ at model initialization for
all agents, and after each generation for individual learner agents. Social
learning agents' $\bar\pi_{ib}$ is initialized as that of its teacher selected
through performance-biased oblique learning.

\vspace{2em}

\begin{table}[h]
    \caption{Dynamic agent-level variables.}
    \label{tab:modelParameters}
    \centering \hspace{-3em}
    \begin{tabular}{cp{4.0in}l} \toprule

        Attribute & Description & Initial value \\

        \midrule

        $s_i$  & Social learner trait: 1 if agent $i$ is social learner; 0 otherwise & 0
        or 1 w/ equal prob. \\

        $\bar\pi_{ib}$ & ``Ledger'': mean payoffs acquired via behavior $b$ by $i$ 
                       & $B$-vector of floating-point zeros \\

        $c_{ib}$ & Count of how many times agent $i$ performed $b$ 
              & $B$-vector of integer zeros \\

        $\pi_i$ & Net payoffs accumulated by $i$ within generation
                                & 0.0 \\
        \bottomrule
    \end{tabular}
\end{table}

\vspace{2em}

@@

@@

In our model uncertainty is a tunable consequence of four environmental features:
%that structure the behavior of agents.  
% Many factors contribute to the 
% uncertainty involved in both social and asocial learning. 
%We model uncertainty in four distinct ways: 
\begin{enumerate}
    \item We vary the frequency at which the optimal behavior changes, which we call \emph{environmental variability}, $u$. At the start of each generation, with probability $u$, a new behavior is assigned a payoff of $\pisub{high}$ while all other behaviors are assigned a payoff of $\pisub{low}$. Otherwise, the same behavior remains optimal across generations. 
    \item We vary the latent expected payoffs yielded by the bandits. For simplicity,
      we assume that in any given environmental state, there is one optimal behavior that
      yields an expected payoff of $\pisub{high}$, while all other behaviors yield a 
      payoff of $\pisub{low} < \pisub{high}$. As $\pilow$ increases towards 
      $\pihigh$, uncertainty in the form of \emph{payoff ambiguity} also increases,
      making it more difficult, and less rewarding, for agents to determine which
      behavior is the optimal one.
      % The difference between these quantities is the \emph{payoff ambiguity}, 
    % \begin{equation}
    %   A_\pi = 1 - (\pisub{high} - \pisub{low}),
    %   \label{eq:payoffAmbiguity}
    % \end{equation}
    % which is maximized when the expected payoffs are close to equal and minimized
    % when they differ greatly. For simplicity we will set $\pisub{high} = 0.9$
    % for all analyses and vary $\pisub{low}$ in order to vary $A_\pi$.
    \item We vary the total \emph{number of available behaviors}, $B$, which is a
      source of uncertainty since agents are less likely to know which behavior yields high payoffs.  
    \item Finally, we vary number of behavioral events per generation, $M$. This can be viewed as the \emph{effective lifespan} of an agent.  Decreasing this lifespan effectively increases the importance of each event for acquiring payoffs. When lifespans are short ($M$ is small), agents experience greater uncertainty about which behavior is optimal given the fewer learning opportunities.
\end{enumerate}

@@

Model dynamics can be split into three parts: initialization,
\emph{intra}generational behavior and payoff accumulation, and
\emph{inter}generational transmission of the social learning trait and oblique
learning (Figure~\ref{fig:IntraInterGenerationalDynamics}). 
The model runs that we analyze here were initialized with $N=100$ agents
to randomly have the social learner trait or not, with no accumulated payoffs, and
blank ``ledgers''. At each time step, agents select one of $B$ possible behaviors
to perform, accumulating a payoff of 1 if the bandit pays off, and 0 if it does
not---a bandit pays off with probability $\pihigh$ if it is the unique optimal
behavior, or with probability $\pilow$ if it is one of $B-1$ non-optimal
behaviors. Agents accumulate payoffs in this way for $M$ time steps in each
generation. At the end of each generation, $N=100$ agents are randomly selected with
replacement to reproduce (asexual, haploid reproduction), 
weighted by their net payoffs over the generation,
i.e., \emph{performance-biased} reproduction. Child agents then learn from 
one teacher, selected again via a form of performance bias: a child agent chooses
its teacher by first selecting a fully random subset of agents from the 
parent generation as potential teachers, then selecting the one with the 
greatest net payoffs among the subset. This process continues until $T=1000$
time steps have elapsed \mt{this should probably be changed depending on $M$ so that
    all analyses use the same number of generations. my informal checks have shown quick
convergence, but i haven't tested b=4 and b=10 as much.}.

@@

\subsubsection{Initialization}

Model initialization includes agent and environment initializations. Agents are
randomly initialized to have the social learning trait or not, with equal 
probability. Agent ``ledgers'' and behavior counts 
are uniformly initialized to the no-knowledge,
blank-slate case where all $\pi_{ib} = 0$ and $c_{ib} = 0$. Both the 
ledger and behavior count vectors have $B$ entries, one for each
behavior the environment affords. Net payoffs for
the generation are also initialized to zero. All agents are initialized with
the same softmax temperature, $\tau$, that guides their behavior selection 
to be either more exploratory (greater $\tau$) or more exploitative of
what agents believe to be the optimal behavior(s) (lesser $\tau$).

The environment is initialized to have $B$ behaviors with one of them selected
at random to be the optimal behavior, denoted $b^*$. Behavior $b^*$ is 
initialized to have a probability $\pihigh$ of paying off 1. All other behaviors
are initialized to have probability $\pilow$ of paying off 1. The modeler must
specify the environmental variability, $u$, and the number of time steps per 
generation, $M$.

\vspace{2em}
\begin{table}[h]
    \caption{Global environmental and cognitive initialization parameters.}
    \label{tab:modelParameters}
    \centering \hspace{-3em}
    \begin{tabular}{cp{4.0in}l} \toprule

        Symbol & Description & Values tested (bold$=$default) \\

        \midrule

        $B$       & Number of possible behaviors (represented by ``bandits'') 
                  & 2, 4, 10 \\

        $\pihigh$ & Probability that the unique optimal behavior pays off 1 
                & \textbf{0.9} \\

        $\pilow$ & Probability one of $B - 1$ non-optimal behaviors pays off 1 
                 & 0.1, 0.45, 0.8 \\

        $\tau$ & Softmax temp.; $\uparrow=$more exploration, $\downarrow=$more
                    exploitation 
               & 0.01, \textbf{0.1}, 1.0 \\

        $u$    & Probability optimal behavior changes between generations 
               & 0.0, 0.1, \ldots, 1.0 \\

        $M$    & Number of time steps per generation & 1, $B/2$, $B$, $2B$, $4B$ \\

        $N_T$    & Number of teachers to pool, from which best selected 
                 & \textbf{5} \mt{TODO: Test more for sensitivity analysis?} \\

        \bottomrule
        \end{tabular} 
\end{table}
\vspace{2em}

\subsubsection{Intra-generational behaviors and payoffs}

Each generation begins with agents initialized either by the $t=0$ initialization
outlined above, or initialized via inter-generational reproduction and learning,
described in detail below. Within generations, there is no social learning.
At each time step within a generation, agents select a behavior to perform
using the softmax algorithm, then update their ledgers $\bar\pi_{ib}$ and behavior
counts $c_{ib}$. If 
the chosen bandit pays off for an agent, its net payoff is incremented by one,
$\pi'_i \leftarrow \pi_i + 1$. This process continues for $M$ time steps, 
when reproduction, learning, and die-off occur, re-initializing the next 
generation for performing $M$ behaviors selected via this same process.

At each time step, 
agent $i$ chooses behavior $b$ at random, with each behavior
weighted by the softmax function applied to that behavior's observed mean payoff
relative to all mean payoffs in the ledger,
\begin{equation}
  \Pr(i \text{ chooses behavior } b) = \frac{\exp(\bar\pi_{ib}
  / \tau) }{ \sum_{b=1}^B \exp(\bar\pi_{ib} / \tau)}.
\end{equation}  
Softmax behavior selection is a
biologically plausible model of behavior~\cite{Schulz2019} 
that enables agents to explore
alternative behaviors sometimes and exploit the best observed behavior other times,
in accordance with Luce's choice axiom~\cite{Luce1959},.
The parameter $\tau$ specifies how frequently alternative behaviors are
explored versus how frequently the best observed behaviors are 
exploited. Greater $\tau$ means more frequent exploration, lesser $\tau$ means
more frequent exploitation. We set $\tau = 0.1$ for
all computational analyses presented in the main paper. We (WILL HAVE) performed
sensitivity analyses and found a weak dependence on $\tau$ that does not affect our
main conclusions.

When agent $i$ performs behavior $b$, $i$'s behavior count 
is incremented by 1, $c'_{ib} \leftarrow c_{ib} + 1$. Agent $i$'s ledger of mean 
payoffs are updated % from $\bar\pi_{ib}$ to $\bar\pi_{ib}'$
using exponential weighted averaging,

\begin{equation}
  \bar\pi_{ib}' = \bar\pi_{ib} +
    \frac{\mathrm{Bandit_{b}(0, 1)} - \bar\pi_{ib}}{c_{ib}'},
\end{equation}
where 
$\mathrm{Bandit}_{b}(0, 1)$
is 0 or 1 depending on the result of the bandit draw for behavior $b$.

\subsubsection{Inter-generational inheritance and social learning}

Every $M$ time steps, agents from the current generation reproduce, teach,
and die, while the next generation to perform the next $M$ time steps learns
from one teacher from the previous generation if they inherited the social 
learner trait. Inter-generational dynamics thus depend on 
two payoff-biased selection mechanisms: reproducer selection
and teacher selection.

$N$ reproducers are sampled from the population with
replacement weighted by net payoffs, i.e., at each of $N$ draws for a parent,
\begin{equation}
  \Pr(\text{Agent $i$ chosen for reproduction}) = \frac{\pi_i}{\sum_{i=1}^N \pi_i}.
\end{equation}
\noindent
We sample with replacement to allow for more successful agents to have multiple
offspring. Child agents inherit their parents' social learning trait, so 
all social learner parents spawn social learner offspring, and parents lacking
the trait spawn offspring lacking the trait.

Child agents with the social learning trait then must select and learn from a
teacher from their parent's generation, including possibly their parent.
Child agents give no preference to whether potential teachers are social 
learners. A child selects a teacher by first selecting a pool of $N_T = 5$ potential
teachers. Among this pool, each child selects the teacher with the greatest
net payoffs. In case of a tie the teacher is chosen at random from those with
the shared maximum net payoff.

Social learner child agents each acquire their teacher's ledger, and all social
learner children behavior counts are reset to 1 to limit ledger values to be between
0 and 1. Child agents without the social learner trait have all ledger and behavior
count values re-initialized to 0. At this point the re-initialized child agents
engage in behavior selection and payoff accumulation described above.

@@

\subsubsection{Outcome variables}

Our primary outcome variable is the mean prevalence of the social learning
trait over $N=100$ agents in 100 trials at the final time step of each trial,
denoted $\meansl$.
Typically at the final time step all agents either have the social learning
trait or do not (see Supplement for a presentation of time series collections for
a representative sampling of \mt{TODO}).

\begin{table}[h]
    \caption{Global environmental and cognitive initialization parameters.}
    \label{tab:modelParameters}
    \centering \hspace{-3em}
    \begin{tabular}{cp{4.0in}l} \toprule

        Symbol & Description & Values \\

        \midrule

        $\meansl$ & Mean social learning prevalence over agents and trials
                  & $\in [0.0, 1.0]$ \\

        $p_{\text{opt}}$ & Fraction of population performing optimal behavior
                         & $\in [0.0, 1.0]$ \\

        \texttt{count}$(b_i)$ & Number of agents performing behavior $b$ 
                              & Behavior-count key-value pairs\\
        \bottomrule
    \end{tabular}
\end{table}

@@

  \item 
    However, more possible behaviors $B$ ($B=2,4,10$ for the first, second, and
    third column respectively in Figure~\ref{fig:results}) can lead
    social learning to be selected for even for large values of $u$, possibly
    due to the low risk and potentially high reward of learning an accurate
    tally of mean payoffs from the previous generation. 
    Since there are many behaviors and
    few observations, softmax weighting will not significantly 
    favor a sub-optimal behavior. However, if one does learn an accurate tally
    then one does have an increased chance of performing the best one.
